The Key to a Data Parallel Compiler

### 2016-04-05 2016-spring-array/2016-array-key.tex
Cleanup the Introduction and Conclusion

Compilers are everywhere, but their design and structure does not scale
well to modern parallel architectures. Traditional compilers are very
difficult to parallelize, and the benefits of modern architectures is
often seen only through the use of distributed build systems which
invoke the single threaded compilers on multiple ``documents'' at
the same time. Future CPU architectures promise to be even more data
parallel, such as the Xeon Phi and the GPGPU architectures. On these
systems it is even harder to obtain good performance for compilers
written in the traditional fashion.

Fundamentally, a compiler is a translator that operates over some sort
of tree and transforms it to another tree of an usually radically
different shape but equivalent ``semantics.'' These algorithms and
transformations show up everywhere, from document processing on the
Web, XML-based publishing workflows, to the programming language
compilers with which most developers are intimately familar.
These transformations differ from other widely studied parallel
graph problems in that they tend to greatly alter the fundamental
structure of the graph/tree, unlike, say, breadth-first search or
single-source shortest path algorithms. This rapidly shifting graph
structure requires different approaches to achieving good parallelism.

There has been some interesting use of GPUs to accelerate certain
compiler analyses and passes.  Unfortunately, attempts to parallelize
compilers usually takes the form of a great deal of work spent
to highly optimize a single compiler pass. The results are often
fragile and require a great deal of effort to maintain and keep in
line with the rest of the compiler. This integration problem makes it
difficult to scale the ``ad hoc'' parallelization effort of individual
compiler passes to a complete system. This is particularly true in
situations where the compiler passes may change rapidly, such as with
XSLT or document publishing, where it is very common for the document
preparer to write or tweak transformation passes to suit the specific
class of documents being handled. In these cases, the expertise and
time involved in parallelizing such transformations makes the ad hoc
approach mostly intractable.

Furthermore, compilers are increasingly becoming a part of a ``live''
and interactive feedback loop.  Just in time compilation and on
the fly static compilation are both used to provide custom binaries
tuned to specific tasks on the fly. These techniques are used in web
services and large scale databases, where in-memory databases may
construct functions on the fly to execute. There is great interest
in moving such systems on to the GPU or other highly data-parallel
architectures, but it would significantly hamper the performance of
the system if the compilation phase of query execution were required
to occur on the CPU, especially at scale where there may be many more
GPU accelerators than CPU resources.

We propose a language-driven approach to compiler construction that
provides a framework for the development of arbitrary compiler
transformations that are data-parallel by construction. These
techniques intentionally optimize for the data-parallel case, which
permits a wide degree of platform independent performance compared to
the traditional approach to compiler design, which requires significant
effort and possibly signficiant transformation of the source code to
achieve good parallel performance on high-parallel architectures.

We use a mathematically oriented, array based language that is
accessible and popular with information scientists. We restrict
this language to permit only the composition of data-flow oriented
array based programs whose core primitives are all data-parallel by
definition. The permitted compositions of these primitives ensures
that the resulting programs are both analyzable and parallel
by construction, permitting a wide array of whole program and
transformation specific optimizations that have been long studied.
The notation is intentially close to mathematical notation, and
as such, the resulting compiler passes more closely resemble the
compact formal descriptions of the transformations rather than complex
recursive algorithms for transforming trees.

At the core of such techniques, we must have some method for
transforming trees in a controlled and predictable manner. We
build on the Nanopass paradigm for compiler construction, wherein
a compiler is developed as a series of small, functional tree
transformations (passes) that have an input AST language and an
output AST language. Through progressive refinement of the tree, the
composition of these passes results in the complete compiler. This is a
highly functional approach, with very little mutation.  Fundamentally,
each pass must identify regions of the tree upon which it works,
do some sort of computation on that sub-tree that has been matched
(often through pattern matching on the nodes) and reassemble the tree
in some new order that matches the desired structure. In other words,
a compiler nanopass can be thought of as an arbitrary computation
over a given selection of sub-trees in an AST, where this computation
could move, relocate, or change the sub-tree.

The key to our language-drive approach is a set of techniques
that enable a data-parallel expression of these arbitrary
transformations. By using a higher-order function called Key, together
with basic array programming techniques and an encoding for the tree
connections called a Node Coordinate matrix, we can identify sub-trees
and select them for computation by their parent-child relationships
in a data-parallel fashion, and then compute on these sub-trees using
traditional array programming techniques.

We have constructed a complete compiler for a superset of our language
called Co-dfns. This compiler is part of a commercial effort to develop
platform independent performance for array programs. The compiler is
developed completely in a data-parallel by construction as described in
this paper, and uses only traditional and well known array primitives
together with function composition to form the entire core compiler
passes and generator. As a result, the compiler takes the form of a
pure data-flow graph without mutation and has an extremely uniform and
concise definition. It uses no recursion, pattern matching, branching,
or other forms of explicit looping. The compiler, together with its
runtime libraries and parser, comes in under 1000 lines of code.

In this paper we fully elucidate three compiler passes from the Co-dfns
compiler to demonstrate the techniques used and we provide further
analysis of how these techniques are used in the other compiler
passes. We further provide a comparative analysis of these passes
against their counterparts implemented using the Nanopass compiler
framework, comparing performance both in terms of programmer effort
and portability/maintainability, as well as runtime performance.

@@

\item We situate this technique into a broader, language-driven
compiler construction strategy that enables the development of parallel
by construction compilers that exhibit desirable programmability
features in terms of code size, complexity, portability, and
maintainability.

\item We provide analysis of these techniques as used in a commercial
compiler project called Co-dfns and report on the results.

\item We compare these techniques to the Nanopass approach to
compiler design in terms of programmer productivity and compiler
runtime performance.

\item We enumerate a set of programming techniques, patterns, and
strategies that complement the Key operator and Node Coordinates
in the construction of tree transformations in general and give some
intuitions on the use of these techniques on arbitrary transformations.
\end{itemize}

@@

We have derived a method of performing computation over sub-trees
selected on the basis of inter-node properties through the use of the
Key (⌸) operator and node coordinates, which enable local computation
of these inter-node properties. This method is both general and direct,
and when combined with traditional and more mundane array programming,
suffices to implement the complete core of a compiler, modulo parsing
and code generation. The method requires no special operations or
unique special casing primitives in the language. Moreover, it is
strictly data-parallel and data-flow, without any complex control flow,
which not only results in concise code, but also enables much easier
analysis of properties such as runtime complexity and verification
of correctness.

>>>>

Compilers represent a peculiar class of tree/graph algorithms that greatly 
alter the underlying structure of the input graph into something very 
different in structure, but equivalent by some measure of interpretation. 
These algorithms are widely applicable, but have stubbornly resisted a 
general approach to parallelization. Most compilers are single-threaded 
or make very limited use of parallelism in an ad hoc way. The design and 
analysis of compilers almost universally deals with compilers as recursive 
traversals over ASTs. Such formulations usually include heavy reliance on 
branching, recursion, and sometimes very sophisticated control flow in 
practice. All of this contributes to make it difficult to effectively 
parallelize such algorithms onto architectures that are sensitive to 
branching and recursion, such as GPGPUs or highly vectorized CPUs.

There has been some success in efficiently executing parts of a compiler 
on the GPU, such as the parser \cite{bunda1984apl}, 
tokenization \cite{bernecky2003tokenizer}, and certain compiler 
analyses \cite{prabhu2011eigencfa,mendez2012inclusion}. However, a generalized 
framework or strategy for parallel compilation remains elusive. One of 
the major difficulties is the recursive and highly branch dependent nature 
of the algorithms. These present challenges that must be overcome in 
vector-centric architectures.

We present a language-driven strategy for creating an inherently parallel 
compiler. We select a set of well known data-parallel primitives, and then 
restrict program construction to the composition of these operations into 
functions, without any forms of non-linear control flow, such as branching, 
recursion, or pattern matching. Programs constructed in such a style are 
data-parallel by construction. A key problem to the construction of compilers 
in such a restricted environment is dealing with the inherently recursive and 
nested structure of the AST. By working over a linearized representation of 
the AST as a matrix, combined with a structure we call a Node Coordinate 
Matrix, and using data-parallel primitives, we are able to tame the recursive 
and nested nature of the AST into something that can be processed handily 
using only data-parallel, data-flow programs, without requiring branching 
or other forms of non-data parallel control flow.

In the following sections we describe the core methods and idioms surrounding 
the AST encoding, construction of the Node Coordinate Matrix, and the use 
of these structures in handling nested, recursive AST relationships. This 
enables us to perform arbitrary computation over arbitrary sub-trees of an 
AST selected by parent-child relationships in the manner often seen in 
compiler transformations. These ideas have been further implemented and tested 
through the implementation of a complete data-parallel by construction 
compiler, called Co-dfns, that compiles a lexically scoped, functionally 
oriented dialect of APL with nested functions. The compiler targets both the 
GPU and the CPU, and its core compiler is implemented in this pure, restricted 
language that uses only function composition over data-parallel primitives.

@@

\item We situation this technique into a broader, language-driven 
strategy for compiler construction that enables the development of parallel 
by construction compilers that are high-level in their implementation and 
exceptionally concise.

\item We provide analysis of these techniques as used in a commercial
compiler project called Co-dfns and report on the results,
including the overall architecture of the compiler, the uses of these 
techniques within the compiler, and the demonstration of these techniques 
applied to two specific compiler passes.

\end{itemize}

@@

We have derived a method of performing computation over sub-trees
selected on the basis of inter-node properties through the use of the
Key (⌸) operator and node coordinates, which enable local computation
of these inter-node properties. This method is both general and direct,
and when combined with traditional and more mundane array programming,
suffices to implement the complete core of a compiler, modulo parsing
and code generation. The method requires no special operations or
unique special casing primitives in the language. Moreover, it is
strictly data-parallel and data-flow, without any complex control flow,
which results in exceptionally concise code. These methods permit a 
general, high-level approach to constructing a compiler that is 
parallel by construction by constricting the language to only a data parallel 
subset of a normal array language.
